New lower bounds for Schur and weak Schur numbers

### 2021-04-08 article/article.tex
Le pouvoir de BibTeX ᕕ( ᐛ )ᕗ

\newtheorem{definition}{Definition}[section]
\newtheorem{notation}[definition]{Notation}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{computational theorem}{Computational Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]

@@

\section{References}

\hypertarget{label1}{\textbf{[1]}} Schur, I. 1917. Uber die Kongruenz \(x^m + y^m = z^m~(\text{mod}~p)\). Jahresbericht der 
Deutschen Mathematikervereinigung 25:114–117.: 114-116. \url{http://eudml.org/doc/145475}

\hypertarget{label2}{\textbf{[2]}} Rowley, F. 2020. New Lower Bounds for Weak Schur Partitions. arXiv 2011.11292, 
\url{https://arxiv.org/pdf/2011.11292.pdf}

\hypertarget{label3}{\textbf{[3]}} Heule, M.J.H. 2017. Schur Number Five. arXiv 1711.08076, \url{https://arxiv.org/pdf/1711.08076.pdf}

\hypertarget{label4}{\textbf{[4]}} S. Eliahou, J.M. Marín, M.P. Revuelta and M.I. Sanz 2012. Weak Schur numbers and the search for G.W. 
Walker’s lost partitions. Computers \& Mathematics with Applications Vol. 63, \url{https://www.sciencedirect.com/science/article/pii/S0898122111009722}

\hypertarget{label5}{\textbf{[5]}} Bierem, A. 2017. CaDiCaL, Lingeling, Plingeling, Treengeling, YalSAT Entering the SAT Competition 2017. 
Proceedinfs of SAT Competition 2017 -- Solver and Benchmark Descriptions \url{http://fmv.jku.at/papers/Biere-SAT-Competition-2017-solvers.pdf}

>>>>

\newtheorem{definition}{Definition}[section]
\newtheorem{notation}[definition]{Notation}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{computational theorem}{Computational Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{remark}{Remark}[section]
\newtheorem{proposition}{Proposition}[section]

@@

\bibliography{biblio}
\bibliographystyle{plain}

### 2021-04-08 article/intro.tex
formatage code
Le pouvoir de BibTeX ᕕ( ᐛ )ᕗ

Schur proved in \hyperlink{label1}{\textbf{[1]}} that given a number of subsets \(n\), there exists a value of \(p\)
such that there exists no partition of \([\![1,q]\!]\) into \(n\) sum-free subsets for any \(q \geqslant p\). A similar
property holds for weakly sum-free subsets (reference necessaire). These observations lead to the following definitions.

\begin{definition}
Let \(n \in \mathbb{N}^*\). There exists a greatest integer that we note \(S(n)\) (\textit{resp. \(WS(n)\)}) such that
\([\![1,S(n)]\!]\) (resp. \([\![1,WS(n)]\!]\)) can be partitioned into \(n\) sum-free subsets (resp. weakly sum-free
subsets). \(S(n)\) is called the \textit{\(n\)\textsuperscript{th} Schur number} and \textit{\(WS(n)\) the \(n\)\textsuperscript{th} weak
Schur number}.
\end{definition}

>>>>

Schur proved in \cite{Schur1917} that given a number of subsets \(n\), there exists a value of \(p\)
such that there exists no partition of \([\![1,q]\!]\) into \(n\) sum-free subsets for any \(q \geqslant p\). A similar
property holds for weakly sum-free subsets (reference necessaire). These observations lead to the following definitions.
\begin{definition}
Let \(n \in \mathbb{N}^*\). There exists a greatest integer that we note \(S(n)\) (\textit{resp. \(WS(n)\)}) such that
\([\![1,S(n)]\!]\) (resp. \([\![1,WS(n)]\!]\)) can be partitioned into \(n\) sum-free subsets (resp. weakly sum-free
subsets). \(S(n)\) is called the \textit{\(n\)\textsuperscript{th} Schur number} and \textit{\(WS(n)\) the
\(n\)\textsuperscript{th} weak
Schur number}.
\end{definition}

### 2021-04-08 article/sat.tex
Le pouvoir de BibTeX ᕕ( ᐛ )ᕗ

We encode the existence of (weakly) sum-free partitions as propositional formulae like in
\hyperlink{label3}{\textbf{[3]}} and then use
SAT solvers to determine whether these formulae are satisfiable.

@@

\subsection{The search space previously used in computer search for lower bounds may not contain the optimal partitions}
Rowley's new lower bound for \(WS(6)\) (642) \hyperlink{label2}{\textbf{[2]}} was a quite significant improvement upon
the former
best known lower bound (582) \hyperlink{label7}{\textbf{[7]}}. This previous lower bound was found using a computer
(often with Monte-Carlo methods) and by making the
assumption that a good partition for \(WS(n+1)\) starts with a good partition for \(WS(n)\) which is true for small
values of \(n\).
Therefore, one may wonder whether the limiting factor are the assumptions or the methods used to search for partitions.
It appears that the search space
induced by these assumptions does not contain the optimal partitions.

@@

This result was obtained in 8 hours with the SAT solver plingeling \hyperlink{label5}{\textbf{[5]}} on a 2.60 GHz Intel
i7 processor PC.
However, simply encoding the existence of such a partition as explained in the previous subsection would not result in a
reasonable
computation time. In order to help the SAT solver, we add additional information in the propositional formula. We did
not quantify the
speedup, but it most likely allowed us to gain several order of magnitude in computation time as we explain in the next
subsection.

@@

\par
\hyperlink{label4}{\textbf{[4]}} verified with a SAT solver that there are no partition in 5 weakly sum-free subsets of
\([\![1,197]\!]\) with
\(A_5 = \{67, 68\} \cup [\![70,134]\!] \cup \{136\}\) in 17 hours and could not provide a similar result when only
assuming \(m_5 = 67\) even after several
weeks of runtime. By using the same method as above, we were able to verify that \(WS(5 | m_5 = 67) = 196\) in 0.5
seconds with the SAT solver glucose \hyperlink{label6}{\textbf{[6]}}
on a 2.60 GHz Intel i7 processor PC (we used the non-parallel version here for the sake of comparison but in the rest of
this subsection, we used the parallel version of glucose).
The additional information we gave to the SAT solver is that every partition of \([\![1,66]\!]\) in 4 weakly sum-free
subsets starts with a partition of
\([\![1,23]\!]\) in 3 weakly sum-free subsets (this can be checked in a few dozens of minutes with a SAT solver). Among
the 3 partitions of \([\![1,23]\!]\) in
3 weakly sum-free subsets, every number always appears in the same subset except for 16 and 17 which can appear in two
different subsets. We hardcoded
this external knowledge in the propositional formula which allowed us to gain several orders of magnitude in computation
time. We also give the stronger
following result.

>>>>

We encode the existence of (weakly) sum-free partitions as propositional formulae like in
\cite{Heule2017} and then use
SAT solvers to determine whether these formulae are satisfiable.

@@

\subsection{The search space previously used in computer search for lower bounds may not contain the optimal partitions}
Rowley's new lower bound for \(WS(6)\) (642) \cite{RowleyWS} was a quite significant improvement upon
the former
best known lower bound (582) \cite{EliahouBook}. This previous lower bound was found using a computer
(often with Monte-Carlo methods) and by making the
assumption that a good partition for \(WS(n+1)\) starts with a good partition for \(WS(n)\) which is true for small
values of \(n\).
Therefore, one may wonder whether the limiting factor are the assumptions or the methods used to search for partitions.
It appears that the search space
induced by these assumptions does not contain the optimal partitions.

@@

This result was obtained in 8 hours with the SAT solver plingeling \cite{Lingeling2017} on a 2.60 GHz Intel
i7 processor PC.
However, simply encoding the existence of such a partition as explained in the previous subsection would not result in a
reasonable
computation time. In order to help the SAT solver, we add additional information in the propositional formula. We did
not quantify the
speedup, but it most likely allowed us to gain several order of magnitude in computation time as we explain in the next
subsection.

@@

\par
\cite{ELIAHOU2012175} verified with a SAT solver that there are no partition in 5 weakly sum-free subsets of
\([\![1,197]\!]\) with
\(A_5 = \{67, 68\} \cup [\![70,134]\!] \cup \{136\}\) in 17 hours and could not provide a similar result when only
assuming \(m_5 = 67\) even after several
weeks of runtime. By using the same method as above, we were able to verify that \(WS(5 | m_5 = 67) = 196\) in 0.5
seconds with the SAT solver glucose \cite{Glucose}
on a 2.60 GHz Intel i7 processor PC (we used the non-parallel version here for the sake of comparison but in the rest of
this subsection, we used the parallel version of glucose).
The additional information we gave to the SAT solver is that every partition of \([\![1,66]\!]\) in 4 weakly sum-free
subsets starts with a partition of
\([\![1,23]\!]\) in 3 weakly sum-free subsets (this can be checked in a few dozens of minutes with a SAT solver). Among
the 3 partitions of \([\![1,23]\!]\) in
3 weakly sum-free subsets, every number always appears in the same subset except for 16 and 17 which can appear in two
different subsets. We hardcoded
this external knowledge in the propositional formula which allowed us to gain several orders of magnitude in computation
time. We also give the stronger
following result.

### 2021-04-08 article/schur.tex
Le pouvoir de BibTeX ᕕ( ᐛ )ᕗ

\qquad In this section, we use Rowley's constructions [7] in a Schur context. To improve lower bounds for Ramsay's
numbers, Rowley
introduces partitions verifying some properties which can be extended using a method which generalizes Abbott-Hanson's
[8] construction.
Rowley named these partitions "templates", and we will keep this name in the entire article. We find suitable templates
and use them
 to find new lower bounds for Schur numbers.

@@

The first inequality comes from the original Schur's paper \hyperlink{label1}{\textbf{[1]}}. The second one are due to
Abott (référence nécessaire)
and the third one to Rowley (reference nécessaire). The other ones are new. \\

@@

\begin{corollary}
The growth rate for Schur (and Ramsey) numbers satisfies \(\gamma \geqslant \sqrt[5]{380} \approx 3.28 \).
\end{corollary}
\textsc{Proof :} It is a mere consequence of the inequality \( S(n+5) \geqslant 380S(n) + 148\). As for Ramsey's
numbers growth rate, a lower bound can be found using Schur's one, thanks to \(S(n) \leqslant R_n(3)-2 \)
(see \hyperlink{label1}{\textbf{[1]}}).

>>>>

\qquad In this section, we use Rowley's constructions \cite{RowleyRamsey} in a Schur context. To improve lower bounds for Ramsay's
numbers, Rowley
introduces partitions verifying some properties which can be extended using a method which generalizes Abbott-Hanson's
\cite{AbbottHanson} construction.
Rowley named these partitions "templates", and we will keep this name in the entire article. We find suitable templates
and use them
 to find new lower bounds for Schur numbers.

@@

The first inequality comes from the original Schur's paper \cite{Schur1917}. The second one are due to
Abott \cite{AbbottHanson}
and the third one to Rowley \cite{RowleyRamsey}. The other ones are new. \\

@@

\begin{corollary}
The growth rate for Schur (and Ramsey) numbers satisfies \(\gamma \geqslant \sqrt[5]{380} \approx 3.28 \).
\end{corollary}
\textsc{Proof :} It is a mere consequence of the inequality \( S(n+5) \geqslant 380S(n) + 148\). As for Ramsey's
numbers growth rate, a lower bound can be found using Schur's one, thanks to \(S(n) \leqslant R_n(3)-2 \)
(see \cite{Schur1917}).
